---
title: "Estimating latent factors: the Kalman filter"
author: "Francesco Zucca, Michael Fehl"
format: pdf
execute:
  echo: false
  warning: false
  message: false
---

# Abstract

# Introduction

In a 1988 working paper, Stock and Watson proposed a way to estimate and forecast the GDP for the US in the short run. The model adopted in their paper is based on the notion that the comovements in many macroeconomic variables that can be captured by a single, underlying, unobserved variable. Therefore, they assumed that the four indicators they use form a dynamic one-factor model with the latent factor that represents economic activity and that, therefore, can be mapped to GDP.

The motive behind this project is to test the model proposed by S&W, verify that the assumptions of one-factor models hold, and test its performance under non-Gaussian distributions.

## 1. The Stock & Watson model

First, we test the model proposed by Stock and Watson. The vector of indicators we will be using is the following:

$$y_t = \begin{bmatrix} y_{1t} \\ y_{2t} \\ y_{3t} \\ y_{4t} \end{bmatrix}$$

where each one of the macroeconomic indicators is represented in log-differences to capture growth rates, and is standardized to adjust for the scale.

The factor we want to extract is the scalar $f_t$, which follows an AR(2) process.

To be able to estimate the factor we apply the Kalman filter to this model, which is defined by two equations:

-   The measurement equation: $y_t = \lambda f_{t} + e_{t}$
-   The transition equation: $f_{t} = \phi f_{t-1} + w_{t}$

The Kalman filter requires errors $e_t$ to be white noise normal. However, to be able to capture business cycles:

$$f_{t} = \phi_{1}f_{t-1} + \phi_{2}f_{t-2} + w_{t}, \quad w_{t} \sim i.i.d. N(0,1)$$

$$e_{it} = \psi_{i1}e_{i,t-1} + \psi_{i2}e_{i,t-2} + \epsilon_{it}, \quad \epsilon_{it} \sim i.i.d. N(0, \sigma_{i}^{2})$$

**add graphical model**

Given that the errors are not white noise, but follow an AR(2) process, they are incorporated into the state vector of the Kalman filter so that the filter's transition matrix is able to track the autocorrelated memory of the error.

**add graphical model**

```{r}
### LIBRARIES

library("tidyverse")
library("readxl")
library("tidyquant")

### LOADING DATA

# IDs
ids <- c(
  "GDPC1",     # Real GDP
  "INDPRO",    # Industrial Production
  "CMRMTSPL",  # Manufacturing & Trade Sales
  "PAYEMS",    # Nonfarm Payrolls
  "W875RX1"    # Real Personal Income less Transfers 
)

# Pulling the data via FRED api
raw_data <- tq_get(ids, get = "economic.data", from = "1969-10-01", to = "2026-01-01")

# Pivoting data in wide format
data_monthly <- raw_data %>%
  pivot_wider(names_from = symbol, values_from = price) %>%
  arrange(date)

# Aligning the GDP measurement
df_aligned <- data_monthly %>%
  arrange(date) %>%
  mutate(GDPC1 = lag(GDPC1, 2))

# Applying log differences to obtain the final dataset
df_clean <- df_aligned %>%
  arrange(date) %>%
  mutate(
    # Log-difference for the 4 monthly variables (lag = 1)
    across(c(INDPRO, CMRMTSPL, PAYEMS, W875RX1), 
           ~ (log(.x) - log(lag(.x, 1))) * 100),
    
    # Log-difference for the quarterly GDP (lag = 3)
    GDPC1 = (log(GDPC1) - log(lag(GDPC1, 3))) * 100
  ) %>%
  # Dropping the very first row since month-over-month lag creates an NA
  drop_na(PAYEMS)

### DEFINING THE DATA ###

# gdp data
gdp <- df_clean %>%
  select(GDPC1) %>%
  drop_na(GDPC1) %>%
  as.matrix()

# Matrix of indicators
y_raw <- df_clean %>%
  select(- date, - GDPC1) %>%
  as.matrix()

### NORMALIZING DATA ###

# demeaning the y time series and standardizing it
y <- scale(y_raw)

TT <- nrow(y) # number of observations
n <- ncol(y) # number of indicators

# normalizing the variance of w to 1
varQ <- 1
```

### 1.1. Verifying the tetrad constraints

The tetrad constraints cannot be identified in a static way anymore, because the one factor model is dynamic. However, we can still verify them on the autocovariance structure (expand the explanation).

```{r}
### DYNAMIC TETRAD CONSTRAINTS ###

# Function to compute empirical cross-covariance matrix at lag tau
cross_cov_matrix <- function(y, tau) {
  TT <- nrow(y)
  # y_t is from tau+1 to TT
  y_t <- y[(tau + 1):TT, ]
  # y_lag is from 1 to TT-tau
  y_lag <- y[1:(TT - tau), ]
  
  # Compute the covariance matrix between y_t and y_lag
  cov_mat <- cov(y_t, y_lag)
  return(cov_mat)
}

# Function to compute the three tetrads for a given covariance matrix
compute_tetrads <- function(sigma) {
  t1 <- sigma[1, 2] * sigma[3, 4] - sigma[1, 3] * sigma[2, 4]
  t2 <- sigma[1, 2] * sigma[3, 4] - sigma[1, 4] * sigma[2, 3]
  t3 <- sigma[1, 3] * sigma[2, 4] - sigma[1, 4] * sigma[2, 3]
  
  return(c(Tetrad_1 = t1, Tetrad_2 = t2, Tetrad_3 = t3))
}

# We want to check this for a few different lags (e.g., lags 1 through 6)
max_lag <- 6
tetrad_results <- matrix(NA, nrow = max_lag, ncol = 3)
colnames(tetrad_results) <- c("Tetrad_1", "Tetrad_2", "Tetrad_3")
rownames(tetrad_results) <- paste("Lag", 1:max_lag)

# Loop through the lags, compute the cross-covariance, and extract tetrads
for (tau in 1:max_lag) {
  sigma_tau <- cross_cov_matrix(y, tau)
  tetrad_results[tau, ] <- compute_tetrads(sigma_tau)
}

# Print the results
print("Empirical Dynamic Tetrad Constraints (Should be close to 0):")
print(round(tetrad_results, 4))

sigma_0 <- cov(y)
tetrad_0 <- compute_tetrads(sigma_0)
print("Lag 0 (contemporaneous):")
print(round(tetrad_0, 4))
```
